Staged Abstract Interpreters

### 2019-04-06 paper/10_intro.tex
Grammar Intro

Roughly at the same time when Futamura projections was envisioned in 1970s,
\citet{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fix-points. However, constructing sound abstract interpreters
was considered abstruse and complicated for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
\cite{DBLP:conf/icfp/HornM10} uncovers a systematic method to derive sound
abstract interpreters from their concrete counterparts, where the soundness can
be easily established by examining the transformation of semantics. The AAM
approach also has been been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.

@@

\begin{itemize}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\item Monadic abstractions, which encapsulate the above two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed.
  %\todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\end{itemize}

@@

%\paragraph{Performance via Specialization}
Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter while the former is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not new, despite the fact that most of the previous work
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation form. In this paper, we advocate a systematic
approach of meta-programming that achieves the performance goal with less
intrusive changes and engineering efforts.

@@

Further more, we show various applications of staging an abstract interpreter.
(1) We compare our approach with abstract compilation, and show that we can
achieve the same optimization by utilizing type-based stage annotations,
meanwhile the analyzer program does not need any changes. (2) \todo{} (3)
Staging abstract interpreter enables modular analysis and provides a mechanized
way to create library summaries. We notice that we may apply the staged abstract
interpreter separately on the different part of the analyzed program, the
generated code is reusable and composable. Therefore, we can turn a
whole-program analyzer into a analyzer that runs modularly.

\iffalse
Our next contribution is to enable modular analysis based on a whole-program
analyzer by meta-programming. Modern software are shipped with large libraries
\cite{toman_et_al:LIPIcs:2017:7121}, and analyzing programs with large libraries
is usually expensive or unnecessarily repeated for whole-program analyzing
algorithms. For example, it has been shown that analyzing a simple ``Hello
World'' program in Java depends on additional 3,000 classes in the library
\cite{DBLP:conf/oopsla/KulkarniMZN16}. A natural idea is to analyze libraries
and generate summaries, later we can plugin the necessary information into the
summaries and finish the analysis; therefore such summaries should be modular
and reusable. Meta-programming gives us a chance to readily tweak a
whole-program analyzing algorithm into an analyzer that works modularly and
produces summaries. Still, the idea is to specialize the abstract interpreter
that was used for whole-program analysis, but on a subset of the program, for
examples, a set of generic functions for list manipulation. Such specialization
works on lambda expressions and generates code, which is partial analysis
results and can be completed by running with the necessary arguments (for
example, an environment that closes the lambda term) at next stages. Multiple
partial analysis results also can be reused later and composed together as they
are just normal programs at the next stage. Through staging, we mechanically
obtain a modular analysis, even though the original algorithm is formulated as a
whole-program analysis.
\fi

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and empirical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while untouch 
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    garbage collection, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive and store-widening analysis.
\end{itemize}

@@

\iffalse
Likewise, specializing static analyses by partial evaluation emerged in late 90s
, and indeed it is able to effectively remove the interpretive
overhead of repeatedly traversing the abstract syntax tree. However, these
previous works focused mostly on one particular analysis, or required to
completely rewrite the analyzer. Hence, it is worth to investigate the idea from
a modern perspective based on generative programming, especially for a general
abstract interpreter that models direct-style $\lambda$-calculus, imperative
features such as mutation, and different abstract domains. One technical
challenge is the binding-time engineering when non-determinism, fixed-point
iterations, different abstract domains and staging are introduced at the same
time. In this paper, we present an end-to-end design and implementation of
staging an abstract interpreter; that means not only the interpreter that
traverses the abstract syntax tree, but also the data structure of abstract
domains, abstract environment and heap, and the fixed-point iteration are
staged.

\todo{rewrite this paragraph}
On the other hand, the abstract interpreter, as a semantic artifact, should be
written in a style that is easy for people to communicate the formulation and
abstraction, but also can be implemented efficiently. As the slogan of
multi-stage programming said, "abstraction without regret", we draw connection
between high-level description and efficient implementation of abstract
interpreters, just like the connection between concrete interpreters and
compilers drawn by Futamura projection. Particularly, we show an easy but
systematic way of adding stage annotations to the abstract interpreter without
changing the code of the interpreter skeleton, which is shared between four concrete
or abstract, unstaged or staged interpreters. We use the LMS framework for staging,
which allows us only use types to annotate the binding-time. Therefore, the
proposed approach bridges the gap between designing sound static analysis and
implementing efficient program analyzer.

@@

When staging a concrete interpreter, the programmers need to distinguish static
and dynamic values --- the given program to be executed by the interpreter is
classified as static because it is known at compile-time, and the inputs to that
program are dynamic. However, when staging an abstract interpreter, this
distinction does not exist anymore. Because the abstract interpreter
instantiates all the inputs as some form of abstract values, which are usually
top elements in their abstract domains and are also statically known. Then what
is the point of staging if there is no such distinction? A surprising by-product
of thinking about this question is to realize that we can apply the staged
abstract interpreter on \textit{open} programs, and the free variables
representing other parts of the program (e.g., libraries) are dynamic inputs,
therefore we obtain a modular analysis through staging, mechanically.
\note{TR: I don't understand this. The program structure is static, the abstract values
are still dynamic, no? They change in every iteration of the fixpoint algorithm}
\note{TR: why is it a big deal. On Closed programs we obtain a constant factor,
but modular analysis has different asymptotics}

@@

It has been observed that partially applying context-sensitivity on selected
portion of the program could improve the precision and efficiency
\cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}. We show that staging
abstract interpreters as an approach to effectively implement hybird
context-sensitivity\todo{}.
\fi

>>>>

Roughly at the same time when Futamura projections were envisioned in the 1970s,
\citet{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fix-points. However, constructing sound abstract interpreters
was considered abstruse and complicated for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
\cite{DBLP:conf/icfp/HornM10} uncovers a systematic method to derive sound
abstract interpreters from their concrete counterparts, where the soundness can
be easily established by examining the transformation of semantics. The AAM
approach also has been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.

@@

\begin{itemize}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores, and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\item Monadic abstractions, which encapsulate the above two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed.
  %\todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\end{itemize}

@@

%\paragraph{Performance via Specialization}
Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter while the former is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not new, despite the fact that most of the previous works
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation form. In this paper, we advocate a systematic
approach of meta-programming that achieves the performance goal with less
intrusive changes and engineering efforts.

@@

Furthermore, we show various applications of staging an abstract interpreter.
(1) We compare our approach with abstract compilation, and show that we can
achieve the same optimization by utilizing type-based stage annotations,
meanwhile, the analyzer program does not need any changes. (2) \todo{} (3)
Staging abstract interpreter enables modular analysis and provides a mechanized
way to create library summaries. We notice that we may apply the staged abstract
interpreter separately on the different part of the analyzed program, the
generated code is reusable and composable. Therefore, we can turn a
whole-program analyzer into an analyzer that runs modularly.

\iffalse
Our next contribution is to enable modular analysis based on a whole-program
analyzer by meta-programming. Modern software are shipped with large libraries
\cite{toman_et_al:LIPIcs:2017:7121}, and analyzing programs with large libraries
is usually expensive or unnecessarily repeated for whole-program analyzing
algorithms. For example, it has been shown that analyzing a simple ``Hello
World'' program in Java depends on additional 3,000 classes in the library
\cite{DBLP:conf/oopsla/KulkarniMZN16}. A natural idea is to analyze libraries
and generate summaries, later we can plug in the necessary information into the
summaries and finish the analysis; therefore such summaries should be modular
and reusable. Meta-programming gives us a chance to readily tweak a
whole-program analyzing algorithm into an analyzer that works modularly and
produces summaries. Still, the idea is to specialize the abstract interpreter
that was used for whole-program analysis but on a subset of the program, for
examples, a set of generic functions for list manipulation. Such specialization
works on lambda expressions and generates code, which is partial analysis
results and can be completed by running with the necessary arguments (for
example, an environment that closes the lambda term) at next stages. Multiple
partial analysis results also can be reused later and composed together as they
are just normal programs at the next stage. Through staging, we mechanically
obtain a modular analysis, even though the original algorithm is formulated as a
whole-program analysis.
\fi

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and empirical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while \note{not a word} untouch 
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    garbage collection, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of the program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive and store-widening analysis.
\end{itemize}

@@

\iffalse
Likewise, specializing static analyses by partial evaluation emerged in the late 90s
, and indeed it is able to effectively remove the interpretive
overhead of repeatedly traversing the abstract syntax tree. However, these
previous works focused mostly on one particular analysis, or required to
completely rewrite the analyzer. Hence, it is worth to investigate the idea from
a modern perspective based on generative programming, especially for a general
abstract interpreter that models direct-style $\lambda$-calculus, imperative
features such as mutation, and different abstract domains. One technical
challenge is the binding-time engineering when non-determinism, fixed-point
iterations, different abstract domains, and staging are introduced at the same
time. In this paper, we present an end-to-end design and implementation of
staging an abstract interpreter; that means not only the interpreter that
traverses the abstract syntax tree, but also the data structure of abstract
domains, abstract environment and heap, and the fixed-point iteration are
staged.

\todo{rewrite this paragraph}
On the other hand, the abstract interpreter, as a semantic artifact, should be
written in a style that is easy for people to communicate the formulation and
abstraction, but also can be implemented efficiently. As the slogan of
multi-stage programming said, "abstraction without regret", we draw a connection
between high-level description and efficient implementation of abstract
interpreters, just like the connection between concrete interpreters and
compilers drawn by the Futamura projections. Particularly, we show an easy but
systematic way of adding stage annotations to the abstract interpreter without
changing the code of the interpreter skeleton, which is shared between four concrete
or abstract, unstaged or staged interpreters. We use the LMS framework for staging,
which allows us only use types to annotate the binding-time. Therefore, the
proposed approach bridges the gap between designing sound static analysis and
implementing efficient program analyzer.

@@

When staging a concrete interpreter, the programmers need to distinguish static
and dynamic values --- the given program to be executed by the interpreter is
classified as static because it is known at compile-time, and the inputs to that
program are dynamic. However, when staging an abstract interpreter, this
distinction does not exist anymore. Because the abstract interpreter
instantiates all the inputs as some form of abstract values, which are usually
top elements in their abstract domains and are also statically known. Then what
is the point of staging if there is no such distinction? A surprising by-product
of thinking about this question is to realize that we can apply the staged
abstract interpreter on \textit{open} programs, and the free variables
representing other parts of the program (e.g., libraries) are dynamic inputs,
therefore we obtain a modular analysis through staging, mechanically.
\note{TR: I don't understand this. The program structure is static, the abstract values
are still dynamic, no? They change in every iteration of the fixpoint algorithm}
\note{TR: why is it a big deal. On Closed programs, we obtain a constant factor,
but modular analysis has different asymptotics}

@@

It has been observed that partially applying context-sensitivity on selected
portions of the program could improve the precision and efficiency
\cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}. We show that staging
abstract interpreters as an approach to effectively implement hybrid
context-sensitivity\todo{}.
\fi
